---
title: "Introduction"
subtitle: "`r paste0('metabolyseR v',packageVersion('metabolyseR'))`"
author: "Jasen Finch"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  prettydoc::html_pretty:
    toc: true
    highlight: github
    theme: tactile
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r,include=FALSE}
knitr::opts_chunk$set(fig.align = 'center')
library(magrittr)
```

## Introduction

The *metabolyseR* package provides a suite of methods that encompass three elements of metabolomics data analysis:

* data pre-treatment
* modelling / data mining
* correlation analyses

The package also distinguishes between the flexibility and simplicity required for **exploratory** analyses compared to the convenience needed for more complex **routine** analyses. 
This is reflected in the underlying S4 object-oriented implementations and associated methods defined within the package.
It should be noted that it is useful to understand the principles involved in using *metabolyseR* for exploratory analyses to aid in extracting and wrangling the results generated from routine analyses.

The following document will provide an introduction to the basic usage of the package and includes how to create and use the base classes that are the foundation of *metabolyseR*.
This will be focused around the applications for both exploratory and routine analyses.
For more detailed information on the individual analysis elements see their associated vignette using:

```{r vignettes, eval=FALSE}
browseVignettes('metabolyseR')
```

There is also an example quick start analysis vignette provided. 

```{r example_analysis,eval=FALSE}
vignette('quick_start','metabolyseR')
```

Any issues, bugs or errors encountered while using the package should be reported [here](https://github.com/jasenfinch/metabolyseR/issues).

The examples shown here will use the `abr1` data set from the [metaboData](https://aberhrml.github.io/metaboData/) package (`?metaboData::abr1`).
This is a nominal mass flow-injection mass spectrometry (FI-MS) fingerprinting data set from a plant-pathogen infection time course experiment. 
The examples will also include use of the pipe `%>%` from the [magrittr](https://magrittr.tidyverse.org/) package.

Firstly load the necessary packages:

```{r libraryLoad,message=FALSE}
library(metabolyseR)
library(metaboData)
```

## Parallel processing

The package supports parallel processing using the [future](https://CRAN.R-project.org/package=future) package.

By default, processing by `metabolyseR` will be done sequentially.
However, parallel processing can be activated, prior to analysis, by specifying a parallel back-end using `plan()`.
The following example specifies using the `multisession` implementation (multiple background R sessions) with two worker processes.

```{r plan}
plan(future::multisession,workers = 2)
```

See the future package [documentation](https://CRAN.R-project.org/package=future) for more information on the types of parallel implementations that are available.

## Exploratory analyses

For exploratory analyses, simple questions of the data need to be answered quickly, requiring few steps.
Key requirements for any tool used by investigators are that it should be both simple and flexible.

In *metabolyseR*, the `AnalysisData` class is the base S4 class that provides these requirements.
The following sections will give an overview of the basics in constructing and using these objects as the base for analysis.

### Analysis data

We can firstly construct an `AnalysisData` object which requires two data tables.
The first is the metabolomic data where the columns are the metabolome features, the rows the sample observations and contains the abundance values.
The second is the sample meta-information where the row order should match to that of the metabolome data table.
Using the example data, his can be constructed and assigned to the variable `d` by:

```{r analysis_data}
d <- analysisData(data = abr1$neg,
                  info = abr1$fact)
```

Where `abr1$neg` is the negative ionisation mode data and `abr1$fact` is the corresponding sample information.
By printing `d` we can view some basic information about our data.

```{r print_analysis_data}
print(d)
```

We can also return the numbers of samples and numbers of features respectively using the following:

```{r number_samples_features}
nSamples(d)
nFeatures(d)
```

The data table can be extracted using the `dat` method:

```{r dat}
dat(d)
```

Or alternatively, can be used to assign a new data table:

```{r new_dat}
dat(d) <- abr1$pos
d
```

The sample information table can be extracted using the `sinfo` method:

```{r sinfo}
sinfo(d)
```

And similarly used to assign a new sample information table:

```{r new_sinfo}
sinfo(d) <- abr1$fact[,1:2]
d
```

```{r analysis_data2,echo=FALSE}
d <- analysisData(abr1$neg,abr1$fact)
```

### Sample information

There are a number of methods that provide utility for querying and altering the sample information within an `AnalysisData` object.
These methods are all named with the prefix `cls` and include:

```{r cls,results='asis',echo=FALSE}
getNamespaceExports('metabolyseR') %>% 
  {.[stringr::str_detect(.,'cls')]} %>% 
  {.[!stringr::str_detect(.,':')]} %>% 
  sort() %>% 
  stringr::str_c('* `',.,'`') %>% 
  stringr::str_c(collapse = '\n') %>%
  cat()
```

The names of the available sample information columns can be shown using `clsAvailable()`.

```{r cls_available}
clsAvailable(d)
```

A given column can be extracted using `clsExtract()`. 
Here, the `day` column is extracted.

```{r cls_extract}
clsExtract(d,cls = 'day')
```

Sample class frequencies could then be computed.

```{r cls_freq}
clsExtract(d,cls = 'day') %>%
  table()
```

It can be seen that there are 20 samples available in each class.

Another example is the addition of a new sample information column.
In the following, a column called `new_class` will be added with all samples labelled `1`.

```{r cls_add}
d <- clsAdd(d,cls = 'new_class',value = rep(1,nSamples(d)))
clsAvailable(d)
```

### Keeping / removing samples or features

Samples or features can easily be kept or removed from an `AnalysisData` object as is most convenient.

Below can be seen the first 6 sample indexes in the `injorder` column of the sample information.

```{r show-samples}
samples <- d %>%
  clsExtract(cls = 'injorder') %>%
  head()

print(samples)
```

Only these samples could be kept using:

```{r keep-samples}
d %>%
  keepSamples(idx = 'injorder',samples = samples)
```

Or removed using:

```{r remove-samples}
d %>%
  removeSamples(idx = 'injorder',samples = samples)
```

The process is very similar for keeping or removing specific metabolome features from the data table.
Below can be seen the first 6 feature names in the data table.

```{r show-features}
feat <- d %>%
  features() %>%
  head()

print(feat)
```

Only these features can be kept using:

```{r keep-features}
d %>%
  keepFeatures(features = feat)
```

Or to remove these features:

```{r remove-features}
d %>%
  removeFeatures(features = feat)
```

## Routine analyses

Routine analyses are those that are often made up of numerous steps where parameters have likely already been previously estabilshed.
The emphasis here is on convenience with as little code as possible required.
In these analyses, the necessary analysis elements, order and parameters are first prepared and then the analysis routine subsequently performed in a single step.
This section will introduce how this type of analysis can be performed using *metabolyseR* and will include four main topics:

* analysis parameter selection
* performing an analysis
* performing a re-analysis
* extracting analysis results

### Analysis parameters

Parameter selection is the fundamental aspect for performing routine analyses using *metabolyseR* and will be the step requiring the most input from the user.
The parameters for an analysis are stored in an S4 object of class `AnalysisParameters` containing the relevant parameters of the selected analysis elements.

The parameters have been named so that they denote the same functionality commonly across all analysis element methods.
**Discussion of the specific parameters can be found withing the vignettes of the relevant analysis elements.**
These can be accessed using:

```{r vignettes1, eval=FALSE}
browseVignettes('metabolyseR')
```

There are several ways to specify the parameters to use for analysis.
The first is programatically and the second is through the use of the YAML format.

#### Programatic specification

The available analysis elements can be shown using:

```{r analysis_elements}
analysisElements()
```

The `analysisParameters()` function can be used to create an `AnalysisParameters` object containing the default parameters.
For example, the code below will return default parameters for all the *metabolyseR* analysis elements.

```{r parametersExample}
p <- analysisParameters()
p
```

To retrieve parameters for a subset of analysis elements the following can be run, returning parameters for only the pre-treatment and modelling elements. 

```{r parametersElementSubset}
p <- analysisParameters(c('pre-treatment','modelling'))
p
```

The `changeParameter()` function can be used to uniformly change these parameters across all of the selected methods.
The example below changes the defaults of all the parameters named `cls` from the default `class` to `day`.

```{r changeParameter}
p <- analysisParameters()
changeParameter(p,'cls') <- 'day'
p
```

Alternatively the parameters of a specific analysis elements can be targeted using the `elements` argument.
The following will only alter the `cls` parameter back to `class` for the pre-treatment element parameters:

```{r changeParameterElement}
changeParameter(p,'cls',elements = 'pre-treatment') <- 'class'
```

Parameters can be extracted from the `AnalysisParameters` class using the `parameters()` function for a specified element.

```{r parameters}
parameters(p,'correlations')
```

Each analysis element has a function for returning default parameters for specific methods.
These include `preTreatmentParameters()`, `modellingParameters()` and `correlationParameters()`.
Each returns a list of the default parameters for a specified methods as shown in the example for `modellingParameters()` below.

```{r modellingParameters}
modellingParameters('anova')
```

Refer to the documentation (`?`) of each function for sepecific usage details.

The parameters returned by these functions can be assigned to an `AnalysisParameters` object, again using `parameters()`'

```{r assign_pre_treatment_parameters}
parameters(p,'pre-treatment') <- preTreatmentParameters(
  list(
    occupancyFilter = 'maximum',
    transform = 'TICnorm'
      )
  )
```

#### YAML specification

Due to the relatively complex structure of the parameters needed for analyses containing many components, it is also possible to specify analysis parameters using the YAML file format.
YAML parameter files (.yaml) can be parsed using the `parseParameters()` function.
The example below shows the YAML specification for the defaults returned by `analysisParameters()`.

```{r defaultYAML,echo=FALSE,results='asis'}
paramFile <- system.file('defaultParameters.yaml',package = 'metabolyseR')

stringr::str_c("
```yaml
",
yaml::read_yaml(paramFile) %>%
  yaml::as.yaml(),
"```") %>%
  cat()
```

This can be passed directly into an `AnalysisParameters` object using the following:

```{r parseParametes}
paramFile <- system.file('defaultParameters.yaml',package = 'metabolyseR')
p <- parseParameters(paramFile)
```

For more complex pre-treatment situations such as the following:

```{r exampleYAML,echo=FALSE,results='asis'}
exampleParamFile <- system.file('exampleParameters.yaml',package = 'metabolyseR')

stringr::str_c("
```yaml
",
  yaml::read_yaml(exampleParamFile) %>%
    yaml::as.yaml(),
  "```") %>%
  cat()
```

Where multiple steps of the same method needed (here is `remove`), these are numbered sequentially.
Where multiple values also need to be provided to a particular argument (e.g. `classes = c('H','1')`), these should be supplied as a hyphenated list.

Existing `AnalysisParameters` objects can also be exported to YAML format as shown below:

```{r export_parameters,eval=FALSE}
p <- analysisParameters()
exportParameters(p,file = 'analysis_parameters.yaml')
```

### Performing an analysis

The analysis is performed in a single step using the `metabolyse()` function.
This accepts the metabolomic data, the sample information and the analysis parameters.

The metabolomic data table of abundance values where the columns are the metabolome features and the rows are each sample observation.
Similarly, the sample meta-information table should consist of the observations as rows and the meta information as columns.
**The order of the observation rows of the sample information table should be concordant with the rows in the metabolomics data table.**

We can run an example analysis using the `abr1` data set by first generating the default parameters for pre-treatment and modelling (random forest) analysis elements.

```{r metabolyse_default_parameters}
p <- analysisParameters(c('pre-treatment','modelling'))
```

Custom pre-treatment parameters can then be specified to only inlude occupancy filtering and total ion count normalisation. 

```{r custom_pre-treatment_parameters}
parameters(p,'pre-treatment') <- preTreatmentParameters(
  list(
  occupancyFilter = 'maximum',
  transform = 'TICnorm')
)
```

Next the `cls` parameters can be changed to use the `day` sample information column throughout the analysis.

```{r change_cls_parameter}
changeParameter(p,'cls') <- 'day'
```

Finally, the analysis can be run in a single step.
Here only the fist 200 features of the negative ionisation mode data are specified to reduce the analysis time needed for this example.

```{r metabolyse}
analysis <- metabolyse(abr1$neg[,1:200],abr1$fact,p) 
```

**Note: If a data pre-treatment step is not performed prior to modelling or correlation analysis, the raw data will automatically be used.**

The `analysis` object containing the analysis results can be printed to provide some basic information about the results of the analysis.

```{r print_analysis_results}
print(analysis)
```

### Performing a re-analysis

There are likely to be occasions where an analysis will need to be re-analysed using a new set of parameters. 
This can be achieved using the `reAnalyse()` function.

In the example below we will run a correlation analysis in addition to the pre-treatment and modelling elements already performed.

Firstly, we can specify the correlation parameters: 

```{r correlation_parameters}
parameters <- analysisParameters('correlations')
```

Then perform the re-analysis on our previously analysed `Analysis` object, specifying the additional parameters.

```{r reAnalyseExample}
analysis <- reAnalyse(analysis,parameters)
```

An overview of the results of the analysis (now including correlations) can then be printed.

```{r print_reanalysis}
print(analysis)
```

### Extracting analysis results

An analysis performed by `metabolyse()` returns an S4 object of class `Analysis`.
There are a number of ways of extracting analysis results from this object.

Similarly to the `AnalysisData` class, the `dat()` and `sinfo()` functions can be used to extract the metabolomics data or sample information tables directly for either the `raw` or `pre-treated` data.

For example, to extract the pre-treated metabolomics data from our object `analysis`:

```{r}
dat(analysis,type = 'pre-treated')
```

Or to extract the raw sample information:

```{r}
sinfo(analysis,type = 'raw')
```

Alternatively the `raw` or `preTreated` functions can be used to extract the `AnalysisData` class objects containing both the metabolomics data and sample information for the raw and pre-treated data respectively.

```{r}
raw(analysis)
```

```{r}
preTreated(analysis)
```

Lastly the `analysisResults` function can be used to extract the results of any of the analysis elements.
The following will extract the modelling results:

```{r}
analysisResults(analysis,element = 'modelling')
```
