---
title: "metabolyseR"
subtitle: "`r paste0('v',packageVersion('metabolyseR'))`"
author: "Jasen Finch"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  prettydoc::html_pretty:
    toc: true
    highlight: github
    theme: tactile
vignette: >
  %\VignetteIndexEntry{Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
  
```{r,include=FALSE}
knitr::opts_chunk$set(fig.align = 'center')
library(magrittr)
```

## Introduction

Extracting biological information, related to experimental treatments, from metabolomics data involves a number of steps; data pre-treatment, assessing the extent of discrimination or association between treatment classes and identifying explanatory features responsible for this.
The *metabolyseR* package provides a suite of methods that encompass these aspects of metabolomics data analysis and include the following three analysis elements:

* data pre-treatment
* modelling / data mining
* correlation analyses

The *metabolyseR* package also distinguishes between the flexibility and simplicity required for **exploratory** analyses and the convenience needed for complex **routine** analyses. 
This is reflected in the underlying S4 classes and associated methods defined within the package.
It should be noted that it is useful to understand the principles involved in using *metabolyseR* for exploratory analyses in manipulating and extracting the results generated from routine analyses.

The following document will provide an introduction to the basic usage of the package that includes how to create an manipulate the base classes that are the foundation of *metabolyseR* .
This will be focused around the application for both exploratory and routine analyses.
For more detailed information on the individual analysis elements see their associated vignette using:

```{r vignettes, eval=FALSE}
browseVignettes('metabolyseR')
```

There is also an example quick start analysis vignette provided. 
This can be found using:

```{r example_analysis,eval=FALSE}
vignette('quick_start','metabolyseR')
```

Any issues or bugs found while using the package should be reported [here](https://github.com/jasenfinch/metabolyseR/issues).

The examples shown here will use the `abr1` data set from the *[metaboData](https://github.com/aberHRML/metaboData)* package (`?metaboData::abr1`).
This is a nominal mass flow-injection mass spectrometry (FI-MS) fingerprinting data set from a plant-pathogen infection time course experiment. 
The examples will also include use of the pipe `%>%` from the [magrittr](https://cran.r-project.org/web/packages/magrittr/index.html) package.

Firstly load the relevant packages:

```{r libraryLoad,message=FALSE}
library(metabolyseR)
library(metaboData)
```

## Exploratory analyses

For exploratory analyses, simple questions of the data need to be answered quickly.
Usually with very few steps required.
This means that key requirements for any tool used by investigators for performing exploratory analyses are both simple and flexible.

In *metabolyseR*, the `AnalysisData` class is the base S4 class that provides these requirements.
This is "requires metabolomics data and sample information, rows observations, columns metabolome features (eg. mass spectral signals)"

The following sections will give an overview of the basics in constructing and manipulating these objects as the base for analysis.

### Analysis data

We can firstly construct an `AnalysisData` object to analyse consisting of two data tables.
The first contains the metabolomic data where the columns are the metabolome features, the rows the sample observations and the values.
The second contains the sample meta-information with the row order matching those of the metabolome data table.
Using the example data, his can be constructed and assigned to the variable `d` by:

```{r analysis_data}
d <- analysisData(data = abr1$neg,
                  info = abr1$fact)
```

Where `abr1$neg` is the negative ionisation mode data and `abr1$fact` is the corresponding sample information.
By printing `d` we can view some basic information about our data.

```{r print_analysis_data}
d
```

We can also return the numbers of samples and numbers of features respectively using the following:

```{r number_samples_features}
nSamples(d)
nFeatures(d)
```

The data table can be extracted using the `dat` method:

```{r dat}
dat(d)
```

Or alternatively, can be used to assign a new data table:

```{r new_dat}
dat(d) <- abr1$pos
d
```

The sample information table can be extracted using the `sinfo` method:

```{r sinfo}
sinfo(d)
```

And similarly used to assign a new sample information table:

```{r new_sinfo}
sinfo(d) <- abr1$fact[,1:2]
d
```

```{r analysis_data2,echo=FALSE}
d <- analysisData(abr1$neg,abr1$fact)
```

### Sample information

There are a number of methods that provide utility for querying and altering the sample information within an `AnalysisData` object.
These methods are all named with the prefix `cls` and include:

```{r cls,results='asis',echo=FALSE}
getNamespaceExports('metabolyseR') %>% 
  {.[stringr::str_detect(.,'cls')]} %>% 
  {.[!stringr::str_detect(.,':')]} %>% 
  sort() %>% 
  stringr::str_c('* `',.,'`') %>% 
  stringr::str_c(collapse = '\n') %>%
  cat()
```

The names of the available sample information columns can be shown using `clsAvailable`.

```{r cls_available}
clsAvailable(d)
```

A given column can be extracted using `clsExtract`. 
Here, the `day` column is extracted.

```{r cls_extract}
clsExtract(d,cls = 'day')
```

Sample class frequencies could then be computed.

```{r cls_freq}
clsExtract(d,cls = 'day') %>%
  table()
```

It cam be seen that there are 20 samples available in each class.

Another example is the addition of a new sample information column.
In the following, a column called `new_class` will be added with all samples labelled `1`.

```{r cls_add}
d <- clsAdd(d,cls = 'new_class',value = rep(1,nSamples(d)))
clsAvailable(d)
```

### Keeping / removing samples or features

Samples or features can easily be kept or removed from an `AnalysisData` object as is most convenient.

Below can be seen the first 6 sample indexes in the `injorder` column of the sample information.

```{r}
samples <- d %>%
  clsExtract(cls = 'injorder') %>%
  head()

print(samples)
```

Only these samples could be kept using:

```{r}
d %>%
  keepSamples(idx = 'injorder',samples = samples)
```

Or removed using:

```{r}
d %>%
  removeSamples(idx = 'injorder',samples = samples)
```

The process is very similar for keeping or removing specific metabolome features from the data table.
Below can be seen the first 6 feature names in the data table.

```{r}
feat <- d %>%
  features() %>%
  head()

print(feat)
```

Only these features can be kept using:

```{r}
d %>%
  keepFeatures(features = feat)
```

Or to remove these features:

```{r}
d %>%
  removeFeatures(features = feat)
```

## Routine analyses

Routine analyses are those that are often made up of numerous steps where parameters have likely already been previously estabilshed.
The emphasis here is on convenience with as little code as possible required.
In these analyses, the necessary analysis elements, order and parameters are first prepared and then the analysis routine subsequently performed in a single step.
This section will introduce how this type of analysis can be performed using *metabolyseR* and will include four main topics:

* analysis parameter selection
* performing an analysis
* performing a reanalysis
* extracting analysis results

### Analysis parameters

Parameter selection is the fundamental aspect for performing routine analyses using *metabolyseR* and will be the step requiring the most input from the user.
The parameters for an analysis are stored in an S4 object of class `AnalysisParameters` containing the relevant parameters of the selected analysis elements.

The parameters have been named so that they denote the same functionality commonly across all analysis element methods.
**Discussion of the specific parameters can be found withing the vignettes of the relevant analysis elements.**
These can be accessed using:

```{r vignettes1, eval=FALSE}
browseVignettes('metabolyseR')
```

There are several ways to specify the parameters to use for analysis.
The first is programatically and the second is through the use of the YAML format.

#### Programatic specification

The available analysis elements can be shown using:

```{r analysis_elements}
analysisElements()
```

The `analysisParameters` function can be used to create an `AnalysisParameters` object containing the default parameters.
For example, the code below will return default parameters for all the *metabolyseR* analysis elements.

```{r parametersExample}
p <- analysisParameters()
p
```

To retrieve parameters for a subset of analysis elements the following can be run, returning parameters for only the pre-treatment and modelling elements. 

```{r parametersElementSubset}
p <- analysisParameters(c('pre-treatment','modelling'))
p
```

The `changeParameter` function can be used to uniformly change these parameters across all of the selected methods.
The example below changes the defaults of all the parameters named "cls" from the default "class" to "day".

```{r changeParameter}
p <- analysisParameters()
changeParameter(p,'cls') <- 'day'
p
```

Alternatively the parameters of a specific analysis elements can be targeted using the "elements" argument.
The following will only alter the "cls" parameter back to "class" for the pre-treatment element parameters:

```{r changeParameterElement}
changeParameter(p,'cls',elements = 'pre-treatment') <- 'class'
```

Parameters can be extracted from the `AnalysisParameters` class using the `parameters` function for a specified element.

```{r parameters}
parameters(p,'correlations')
```

Each analysis element has a function for returning default parameters for specific methods.
These include `preTreatmentParameters`, `modellingParameters` and `correlationParameters`.
Each returns a list of the default parameters for a specified methods as shown in the example for `modellingParameters` below.

```{r modellingParameters}
modellingParameters('anova')
```

Refer to the documentation (`?`) of each function for sepecific usage details.

The parameters returned by these functions can be assigned to an `AnalysisParameters` object, again using `parameters`'

```{r assign_pre_treatment_parameters}
parameters(p,'pre-treatment') <- preTreatmentParameters(
  list(
    occupancyFilter = 'maximum',
    transform = 'TICnorm'
      )
  )
```

#### YAML specification

Due to the relatively complex structure of the parameters needed for the analyses containing many components, it is also possible to specify analysis parameters using the YAML format.
YAML parameter files (.yaml) can be parsed using the `parseParameters` function.
The example below shows the YAML specification for the defaults returned by `analysisParameters`.

```{r defaultYAML,echo=FALSE,results='asis'}
paramFile <- system.file('defaultParameters.yaml',package = 'metabolyseR')

stringr::str_c("
```yaml
",
yaml::read_yaml(paramFile) %>%
  yaml::as.yaml(),
"```") %>%
  cat()
```

This can be passed directly into an `AnalysisParameters` object using the following:

```{r parseParametes}
paramFile <- system.file('defaultParameters.yaml',package = 'metabolyseR')
p <- parseParameters(paramFile)
```

For more complex pre-treatment situations such as the following:

```{r exampleYAML,echo=FALSE,results='asis'}
exampleParamFile <- system.file('exampleParameters.yaml',package = 'metabolyseR')

stringr::str_c("
```yaml
",
  yaml::read_yaml(exampleParamFile) %>%
    yaml::as.yaml(),
  "```") %>%
  cat()
```

Where multiple multiple steps of the same method needed (here is "remove"), these are numbered sequentially.
Also where multiple values need to be provided to a particular argument (e.g. `classes = c('H','1')`), these should be supplied as a hyphenated list.

Existing `AnalysisParameters` objects can also be exported to YAML format as shown below:

```{r export_parameters,eval=FALSE}
p <- analysisParameters()
exportParameters(p,file = 'analysis_parameters.yaml')
```


### Performing an analysis

Executing analyses is simple using the `metabolyse` function with specification of the data to use, the sample meta-information (info) and the analysis parameters.
This is shown by the following code.

```{r metabolyseExample}
p <- analysisParameters(c('pre-treatment','modelling'))

parameters(p,'pre-treatment') <- preTreatmentParameters(
  list(
  occupancyFilter = 'maximum',
  transform = 'TICnorm')
)

analysis <- metabolyse(abr1$neg[,1:200],abr1$fact,p) 

print(analysis)
```

The data and info can be either `tibble` or `data.frame` structures and the parameters should be an S4 object of class `AnalysisParameters`.
The order of the rows of the info table should agree with those in the data table.

If data pre-treatment has not been performed prior to modelling, the raw data will automatically be used.
The date and time of each of analysis of each of the elements is also stored and can is displayed when the object is printed.

### Performing a reanalysis

There are likely to be occasions where analyses need to be re-analysed using a new set of parameters. 
`reAnalyse` can be used for this.
It simply requires the specification of the `Analysis` object and parameters for the re-analysis as shown below.

```{r reAnalyseExample,eval=FALSE}
parameters <- analysisParameters('correlations')
analysis <- reAnalyse(analysis,parameters)
analysis
```

### Extracting analysis results


`metabolyse` returns an S4 object of class `Analysis`.
Data and analysis results can be extracted from this object using the following functions:

* `rawData()` - raw data.
* `rawInfo()` - raw info.
* `preTreatedData()` - pre-treated data.
* `preTreatedInfo()` - pre-treated info.
* `modellingResults()` - modelling results.
* `correlationResults()` - correlation results.