---
title: "metabolyseR"
author: "Jasen Finch"
date: "July 2017"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{metabolyseR-usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r libraryLoad,echo=FALSE}
suppressPackageStartupMessages(library(metabolyseR))
```

## Introduction

Extracting biological information  related to experimental treatments from metabolomics data involves a number of steps; data pre-treatment, assessing the extent discrimination between treatment classes and identifying explanatory features responsible for this discrimination.
The metabolyseR oackage provides a suite of methods that encompass these aspects of metabolomics data analysis.
Below shows the four analysis elements in that make up metabolyseR.

```{r metabolyseRDiagram,echo=FALSE,fig.width=6}
  DiagrammeR::grViz(readr::read_file('figures/metabolyseR.gv'))
```

This document will provide a tutorial on the basic usage of the package, followed by more detailed discussion of each of the analysis elements.
Lastly, there will be discussion of the tools for the visualisation of the results from these data as well as the integration of metabolyseR into wider metabolomics workflows.
This allows more bespoke analyses to be performed based on the particular metabolomics technique being used.

## Basic Usage

```{r parametersExample,eval=FALSE}
parameters <- analysisParameters()
parameters
```

```{r metabolyseExample,eval=FALSE}
data(abr1,package = 'FIEmspro')
parameters <- analysisParameters()
parameters@preTreat <- list(
    occupancyFilter = list(maximum = list()),
    transform = list(TICnorm = list())
)
analysis <- metabolyse(abr1$neg,abr1$fact,parameters) 
```

```{r reAnalyseExample,eval=FALSE}
parameters <- analysisParameters('classification')
analysis <- reAnalyse(analysis,parameters)
```

## Pre-Treatment

```{r preTreatmentDiagram,echo=FALSE,fig.width=6}
  DiagrammeR::grViz(readr::read_file('figures/preTreat.gv'))
```

```{r preTreatExample}
data(abr1,package = 'FIEmspro')
parameters <- analysisParameters('preTreat')
parameters@preTreat <- list(
    remove = list(
      class = list(cls = 'class',
                               classes = unique(abr1$fact$class)[!(unique(abr1$fact$class) %in% c(1,2))])),
    occupancyFilter = list(
      maximum = list()),
    transform = list(
      TICnorm = list())
)
analysis <- metabolyse(abr1$neg,abr1$fact,parameters)
```

### Remove Samples and Classes
```{r removeMethodsTable,echo=FALSE,results='asis'}
desc <- metabolyseR:::removeMethods(description = T)
args <- lapply(desc,function(x){return(x$arguments)})
args <- lapply(args,function(x){tibble::tibble(Parameter = names(x),Value = unlist(x))})
args <- lapply(args,function(x){apply(x,1,function(y){paste(y,collapse = ' - ')})})
if (T %in% (sapply(args,length) > 1)) {
  args[sapply(args,length) > 1] <- lapply(args[sapply(args,length) > 1],function(x){paste(x,collapse = '; ')})
}
args <- unlist(args)
desc <- unlist(lapply(desc,function(x){return(x$description)}))
meth <- names(metabolyseR:::removeMethods())
defaults <- lapply(metabolyseR:::removeMethods(),formals)
defaults <- lapply(defaults,function(x){x$dat <- NULL;return(x)})
defaults <- lapply(defaults,function(x){tibble::tibble(Parameter = names(x),Value = unlist(x))})
defaults <- lapply(defaults,function(x){apply(x,1,function(y){paste(y,collapse = ' = ')})})
if (T %in% (sapply(defaults,length) > 1)) {
  defaults[sapply(defaults,length) > 1] <- lapply(defaults[sapply(defaults,length) > 1],function(x){paste(x,collapse = '; ')})
}
defaults <- unlist(defaults)
tab <- tibble::tibble(Method = meth, Description = desc, Arguments = args, Defaults = defaults)

knitr::kable(tab)
```

### Filter Variables based on occupancy
```{r occupancyMethodsTable,echo=FALSE,results='asis'}
desc <- metabolyseR:::occupancyMethods(description = T)
args <- lapply(desc,function(x){return(x$arguments)})
args <- lapply(args,function(x){tibble::tibble(Parameter = names(x),Value = unlist(x))})
args <- lapply(args,function(x){apply(x,1,function(y){paste(y,collapse = ' - ')})})
if (T %in% (sapply(args,length) > 1)) {
  args[sapply(args,length) > 1] <- lapply(args[sapply(args,length) > 1],function(x){paste(x,collapse = '; ')})
}
args <- unlist(args)
desc <- unlist(lapply(desc,function(x){return(x$description)}))
meth <- names(metabolyseR:::occupancyMethods())
defaults <- lapply(metabolyseR:::occupancyMethods(),formals)
defaults <- lapply(defaults,function(x){x$dat <- NULL;return(x)})
defaults <- lapply(defaults,function(x){tibble::tibble(Parameter = names(x),Value = unlist(x))})
defaults <- lapply(defaults,function(x){apply(x,1,function(y){paste(y,collapse = ' = ')})})
if (T %in% (sapply(defaults,length) > 1)) {
  defaults[sapply(defaults,length) > 1] <- lapply(defaults[sapply(defaults,length) > 1],function(x){paste(x,collapse = '; ')})
}
defaults <- unlist(defaults)
tab <- tibble::tibble(Method = meth, Description = desc, Arguments = args, Defaults = defaults)

knitr::kable(tab)
```

### Filter Variables based on QC samples
```{r QCMethodsTable,echo=FALSE,results='asis'}
desc <- metabolyseR:::QCMethods(description = T)
args <- lapply(desc,function(x){return(x$arguments)})
args <- lapply(args,function(x){tibble::tibble(Parameter = names(x),Value = unlist(x))})
args <- lapply(args,function(x){apply(x,1,function(y){paste(y,collapse = ' - ')})})
if (T %in% (sapply(args,length) > 1)) {
  args[sapply(args,length) > 1] <- lapply(args[sapply(args,length) > 1],function(x){paste(x,collapse = '; ')})
}
args <- unlist(args)
desc <- unlist(lapply(desc,function(x){return(x$description)}))
meth <- names(metabolyseR:::QCMethods())
defaults <- lapply(metabolyseR:::QCMethods(),formals)
defaults <- lapply(defaults,function(x){x$dat <- NULL;return(x)})
defaults <- lapply(defaults,function(x){tibble::tibble(Parameter = names(x),Value = unlist(x))})
defaults <- lapply(defaults,function(x){apply(x,1,function(y){paste(y,collapse = ' = ')})})
if (T %in% (sapply(defaults,length) > 1)) {
  defaults[sapply(defaults,length) > 1] <- lapply(defaults[sapply(defaults,length) > 1],function(x){paste(x,collapse = '; ')})
}
defaults <- unlist(defaults)
tab <- tibble::tibble(Method = meth, Description = desc, Arguments = args, Defaults = defaults)

knitr::kable(tab)
```

### Impute missing data
```{r imputeMethodsTable,echo=FALSE,results='asis'}
desc <- metabolyseR:::imputeMethods(description = T)
args <- lapply(desc,function(x){return(x$arguments)})
args <- lapply(args,function(x){tibble::tibble(Parameter = names(x),Value = unlist(x))})
args <- lapply(args,function(x){apply(x,1,function(y){paste(y,collapse = ' - ')})})
if (T %in% (sapply(args,length) > 1)) {
  args[sapply(args,length) > 1] <- lapply(args[sapply(args,length) > 1],function(x){paste(x,collapse = '; ')})
}
args <- unlist(args)
desc <- unlist(lapply(desc,function(x){return(x$description)}))
meth <- names(metabolyseR:::imputeMethods())
defaults <- lapply(metabolyseR:::imputeMethods(),formals)
defaults <- lapply(defaults,function(x){x$dat <- NULL;return(x)})
defaults <- lapply(defaults,function(x){tibble::tibble(Parameter = names(x),Value = unlist(x))})
defaults <- lapply(defaults,function(x){apply(x,1,function(y){paste(y,collapse = ' = ')})})
if (T %in% (sapply(defaults,length) > 1)) {
  defaults[sapply(defaults,length) > 1] <- lapply(defaults[sapply(defaults,length) > 1],function(x){paste(x,collapse = '; ')})
}
defaults <- unlist(defaults)
tab <- tibble::tibble(Method = meth, Description = desc, Arguments = args, Defaults = defaults)

knitr::kable(tab)
```

### Transform the data
```{r transformMethods,echo=FALSE,results='asis'}
desc <- metabolyseR:::transformMethods(description = T)
args <- lapply(desc,function(x){return(x$arguments)})
args <- lapply(args,function(x){tibble::tibble(Parameter = names(x),Value = unlist(x))})
args <- lapply(args,function(x){apply(x,1,function(y){paste(y,collapse = ' - ')})})
if (T %in% (sapply(args,length) > 1)) {
  args[sapply(args,length) > 1] <- lapply(args[sapply(args,length) > 1],function(x){paste(x,collapse = '; ')})
}
args <- unlist(args)
if (T %in% sapply(args,function(x){grepl("''",x)})) {
  args[sapply(args,function(x){grepl("''",x)})] <- lapply(args[sapply(args,function(x){grepl("''",x)})],function(x){''})
}
desc <- unlist(lapply(desc,function(x){return(x$description)}))
meth <- names(metabolyseR:::transformMethods())
defaults <- lapply(metabolyseR:::transformMethods(),formals)
defaults <- lapply(defaults,function(x){x$dat <- NULL;return(x)})
if (T %in% sapply(defaults,is.null)) {
  defaults[sapply(defaults,is.null)] <- lapply(defaults[sapply(defaults,is.null)],function(x){c(`''` = '')})
}
defaults <- lapply(defaults,function(x){tibble::tibble(Parameter = names(x),Value = unlist(x))})
defaults <- lapply(defaults,function(x){apply(x,1,function(y){paste(y,collapse = ' = ')})})

if (T %in% sapply(defaults,function(x){grepl("''",x)})) {
  defaults[sapply(defaults,function(x){grepl("''",x)})] <- lapply(defaults[sapply(defaults,function(x){grepl("''",x)})],function(x){''})
}

if (T %in% (sapply(defaults,length) > 1)) {
  defaults[sapply(defaults,length) > 1] <- lapply(defaults[sapply(defaults,length) > 1],function(x){paste(x,collapse = '; ')})
}
defaults <- unlist(defaults)
tab <- tibble::tibble(Method = meth, Description = desc, Arguments = args, Defaults = defaults)

knitr::kable(tab)
```

## Classification

```{r classificationParameters}
parameters <- analysisParameters('classification')
parameters
```

```{r classificationAnalysis,eval=FALSE}
analysis <- reAnalyse(analysis,parameters)
```

## Feature Selection

```{r featureSelectionParameters}
parameters <- analysisParameters('featureSelection')
parameters
```

```{r featureSelectionExample,eval=FALSE}
analysis <- reAnalyse(analysis,parameters)
```

```{r fsMethodsTable,echo=FALSE,results='asis'}
desc <- metabolyseR:::fsMethods(description = T)
args <- lapply(desc,function(x){return(x$arguments)})
args <- lapply(args,function(x){tibble::tibble(Parameter = names(x),Value = unlist(x))})
args <- lapply(args,function(x){apply(x,1,function(y){paste(y,collapse = ' - ')})})
if (T %in% (sapply(args,length) > 1)) {
  args[sapply(args,length) > 1] <- lapply(args[sapply(args,length) > 1],function(x){paste(x,collapse = '; ')})
}
args <- unlist(args)
desc <- unlist(lapply(desc,function(x){return(x$description)}))
meth <- names(metabolyseR:::fsMethods())
defaults <- lapply(metabolyseR:::fsMethods(),formals)
defaults <- lapply(defaults,function(x){x$dat <- NULL;return(x)})
defaults <- lapply(defaults,function(x){tibble::tibble(Parameter = names(x),Value = unlist(x))})
defaults <- lapply(defaults,function(x){apply(x,1,function(y){paste(y,collapse = ' = ')})})
if (T %in% (sapply(defaults,length) > 1)) {
  defaults[sapply(defaults,length) > 1] <- lapply(defaults[sapply(defaults,length) > 1],function(x){paste(x,collapse = '; ')})
}
defaults <- unlist(defaults)
tab <- tibble::tibble(Method = meth, Description = desc, Arguments = args, Defaults = defaults)

knitr::kable(tab)
```

## Correlations

```{r correlationsParameters}
parameters <- analysisParameters('correlations')
parameters
```

```{r correlationsExample,eval=FALSE}
analysis <- reAnalyse(analysis,parameters)
```

## Visualisation and Workflows
